
<!-- badges: start -->
[![R-CMD-check](https://github.com/reconhub/epitrix/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/reconhub/epitrix/actions/workflows/R-CMD-check.yaml)
[![codecov.io](https://codecov.io/github/reconhub/epitrix/coverage.svg?branch=master)](https://codecov.io/github/reconhub/epitrix?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/epitrix)](https://cran.r-project.org/package=epitrix)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/epitrix)](https://cran.r-project.org/package=epitrix)
[![Downloads from Rstudio mirror](https://cranlogs.r-pkg.org/badges/grand-total/epitrix)](http://www.r-pkg.org/pkg/epitrix)
[![Codecov test coverage](https://codecov.io/gh/reconhub/epitrix/branch/master/graph/badge.svg)](https://app.codecov.io/gh/reconhub/epitrix?branch=master)
<!-- badges: end -->




# Welcome to the *epitrix* package!

This package implements small helper functions usefull in infectious disease
modelling and epidemics analysis.


## Installing the package

To install the current stable, CRAN version of the package, type:
```{r install, eval = FALSE}
install.packages("epitrix")
```

To benefit from the latest features and bug fixes, install the development,
*github* version of the package using:

```{r install2, eval = FALSE}
devtools::install_github("reconhub/epitrix")
```

Note that this requires the package *devtools* installed.





# What does it do?

The main features of the package include:

- **`gamma_shapescale2mucv`**: convert shape and scale of a Gamma distribution
    to mean and CV

- **`gamma_mucv2shapescale`**: convert mean and CV of a Gamma distribution to
    shape and scale

- **`gamma_log_likelihood`**: Gamma log-likelihood using mean and CV

- **`r2R0`**: convert growth rate into a reproduction number

- **`lm2R0_sample`**: generates a distribution of R0 from a log-incidence linear
    model

- **`fit_disc_gamma`**: fits a discretised Gamma distribution to data (typically
    useful for describing delays)

- **`clean_labels`**: generate portable labels by removing non-standard characters or
  replacing them with their closest alphanumeric matches, standardising
  separators, etc.

- **`hash_names`**: generate unique, anonymised, reproducible labels from
    various data fields (e.g. First name, Last name, Date of birth).

- **`emperical_incubation_dist()`** will estimate the empirical incubation
  distribution if given a data frame with dates of onset and a range of
  exposure dates.

- **`fit_gamma_incubation_dist()`** wraps `empirical_incubation_dist()` and
  `fit_disc_gamma()` to fit a discretized gamma distribution to an empirical
  incubation distribution


# Resources

## Worked examples


### Fitting a gamma distribution to delay data

In this example, we simulate data which replicate the serial interval (SI),
i.e. the delays between primary and secondary symptom onsets, in Ebola Virus
Disease (EVD). We start by converting previously estimates of the mean and
standard deviation of the SI (WHO Ebola Response Team (2014) NEJM 371:1481–1495)
to the parameters of a Gamma distribution:

```{r generate_data}
library(epitrix)

mu <- 15.3 # mean in days days
sigma <- 9.3 # standard deviation in days
cv <- sigma / mu # coefficient of variation
cv
param <- gamma_mucv2shapescale(mu, cv) # convertion to Gamma parameters
param

```

The *shape* and *scale* are parameters of a Gamma distribution we can use to
generate delays. However, delays are typically reported per days, which implies
a discretisation (from continuous time to discrete numbers). We use the package
[*distcrete*](https://github.com/reconhub/distcrete) to achieve this
discretisation. It generates a list of functions, including one to simulate data
(`$r`), which we use to simulate 500 delays:

```{r si}

si <- distcrete::distcrete("gamma", interval = 1,
               shape = param$shape,
               scale = param$scale, w = 0)
si
set.seed(1)
x <- si$r(500)
head(x, 10)
hist(x, col = "grey", border = "white",
     xlab = "Days between primary and secondary onset",
     main = "Simulated serial intervals")

```

`x` contains simulated data, for illustrative purpose. In practice, one would
use real data from an ongoing outbreaks. Now we use `fit_disc_gamma` to estimate
the parameters of a dicretised Gamma distribution from the data:

```{r fit}

si_fit <- fit_disc_gamma(x)
si_fit

```


### Converting a growth rate (r) to a reproduction number (R0)

The package [*incidence*](https://github.com/reconhub/incidence) can fit a
log-linear model to incidence curves (function `fit`), which produces a growth
rate (r). This growth rate can in turn be translated into a basic reproduction
number (R0) using `r2R0`. We illustrate this using simulated Ebola data from the
[*outbreaks*](https://github.com/reconhub/outbreaks) package, and using the
serial interval from the previous example:

```{r fit_i}

library(outbreaks)
library(incidence)

i <- incidence(ebola_sim$linelist$date_of_onset)
i
f <- fit(i[1:150]) # fit on first 150 days
plot(i[1:200], fit = f, color = "#9fc2fc")

r2R0(f$info$r, si$d(1:100))
r2R0(f$info$r.conf, si$d(1:100))

```

In addition, we can also use the function `lm2R0_sample` to generate samples of
R0 values compatible with a model fit:

```{r sample_R0}

R0_val <- lm2R0_sample(f$model, si$d(1:100), n = 100)
head(R0_val)
hist(R0_val, col = "grey", border = "white")

```


### Standardising labels

If you want to use labels that will work across different computers, independent
of local encoding and operating systems, `clean_labels` will make your life
easier. The function transforms character strings by replacing diacritic symbols
with their closest alphanumeric matches, setting all characters to lower case,
and replacing various separators with a single, consistent one.


For instance:
```{r clean_labels}
x <- " Thîs- is A   wêïrD LäBeL .."
x
clean_labels(x)

variables <- c("Date.of.ONSET ",
               "/  date of hôspitalisation  /",
               "-DäTÈ--OF___DîSCHARGE-",
               "GEndèr/",
               "  Location. ")
variables
clean_labels(variables)

```

### Anonymising data

`hash_names` can be used to generate hashed labels from linelist data. Based on
pre-defined fields, it will generate anonymous labels. This system has the
following desirable features:

- given the same input, the output will always be the same, so this encoding
  system generates labels which can be used by different people and
  organisations

- given different inputs, the output will always be different; even minor
  differences in input will result in entirely different outputs

- given an output, it is very hard to infer the input (it requires hacking
  skills); if security is challenged, the hashing algorithm can be 'salted' to
  strengthen security

```{r}

first_name <- c("Jane", "Joe", "Raoul", "Raoul")
last_name <- c("Doe", "Smith", "Dupont", "Dupond")
age <- c(25, 69, 36, 36)

## detailed output by default
hash_names(first_name, last_name, age)

## short labels for practical use
hash_names(first_name, last_name, age,
           size = 8, full = FALSE)

```

### Estimate incubation periods

The function `empirical_incubation_dist()` computes the discrete probability distribution by giving equal weight to each patient. Thus, in the case of `N` patients, the `n` possible exposure dates of a given patient get the overall weight `1/(n*N)`. The function returns a data frame with column `incubation_period` containing the different incubation periods with a time step of one day and their `relative_frequency`.

Load environment:

```{r, echo = TRUE, message=FALSE}
library(magrittr)
library(linelist)
library(epitrix)
library(distcrete)
library(ggplot2)
```

Make a linelist object containing toy data with several possible exposure dates for each case:

```{r, echo = TRUE}
ll <- messy_data() %>% clean_data()

x <- 0:15
y <- distcrete("gamma", 1, shape = 12, rate = 3, w = 0)$d(x)
mkexposures <- function(i) i - sample(x, size = sample.int(5, size = 1), replace = FALSE, prob = y)
exposures <- sapply(ll$date_of_onset, mkexposures)
ll$dates_exposure <- exposures

print(ll)
```

Empirical distribution:

```{r}
incubation_period_dist <- empirical_incubation_dist(ll, date_of_onset, dates_exposure)
print(incubation_period_dist)

ggplot(incubation_period_dist, aes(incubation_period, relative_frequency)) +
  geom_col()
```

Fit discrete gamma:

```{r}
fit <- fit_gamma_incubation_dist(ll, date_of_onset, dates_exposure)
print(fit)

x = c(0:10)
y = fit$distribution$d(x)
ggplot(data.frame(x = x, y = y), aes(x, y)) +
  geom_col(data = incubation_period_dist, aes(incubation_period, relative_frequency)) +
  geom_point(stat="identity", col = "red", size = 3) +
  geom_line(stat="identity", col = "red")
```

**Note** that if the possible exposure dates are consecutive for all patients then `empirical_incubation_dist()` and `fit_gamma_incubation_dist()` can take date ranges as inputs instead of lists of individual exposure dates (see help for details).


## Vignettes

The [overview vignette](http://www.repidemicsconsortium.org/epitrix/articles/epitrix.html)
essentially replicates the content of this `README`. To request or contribute
other vignettes, see the section "*getting help, contributing*".

The [estimate incubation vignette](http://www.repidemicsconsortium.org/epitrix/articles/estimate_incubation.html) contains worked examples for the `emperical_incubation_dist()`
`fit_gamma_incubation_dist()`.


## Websites

Click [here](http://www.repidemicsconsortium.org/epitrix/) for the website
dedicated to *epitrix*.


## Getting help, contributing

Bug reports and feature requests should be posted on *github* using the
[*issue*](http://github.com/reconhub/epitrix/issues) system. All other questions
should be posted on the [**RECON forum**](http://www.repidemicsconsortium.org/forum/).

Contributions are welcome via **pull requests**.

Please note that this project is released with a
[Contributor Code of Conduct](CONDUCT.md).
By participating in this project you agree to abide by its terms.
